\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Edit

An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space.
In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
\[
\begin{aligned}
% \log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
% \log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\log (n) &= \eta_1,\\
\log (w) &= \eta_2.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

The design goal of \sdmTMB\ was to develop a flexible, modular, and familiar inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable far beyond ecology.

@@

>>>>

An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}.
The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space and the partial effect of a coefficient from both linear predictors can be combined.
In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
\[
\begin{aligned}
% \log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
% \log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\log (n) &= \eta_1,\\
\log (w) &= \eta_2.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$.
These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$.
These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

The design goal of \sdmTMB\ was to develop a flexible, modular, and familiar inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.

@@

\texttt{summary()} prints standard model information:
